VariantSpark: Population Scale Clustering of Genotype Information

### 2015-08-17 doc/publications/journal2014/paper.tex

Clustering the samples 30\% fastest in \variantSpark\ compared to ADAM (1min and 20secs compared to 1min and 52secs) and  %80 vs 112
90\% faster than Hadoop (14mins and 23secs), see Table~\ref{fivewaycomparison}.
The similarity between \variantSpark\ and ADAM is expected, as they both use {\sc Spark}'s MLlib k-means implementation. 
The 30 second speed increase in \variantSpark\ is likely due us converting the VCF files to sparse vectors, whereas ADAM creates dense vectors.
%Due to the sparse nature of these row-vectors, the sparse vector format is likely to reduce some of the initial overhead.
We also compared the k-means clustering times to a multi-threaded approach in Python and R. 
Although an order of magnitude slower than \variantSpark\ and ADAM, k-means clustering in both Python and R were faster than our Hadoop implementation. Python required 11min and 29sec, whereas R required 7min and 25sec. 
This demonstrates one of Hadoop's disadvantages

 (one which {\sc Spark} dramatically improves on). Hadoop can scale to datasets that exceed memory size as wall as distributing work across nodes in a cluster. However, in this example, the entire dataset can fit in the memory on a single machine. The k-means implemtations
we use in R and Python cluster the in-memory row vectors. Hadoop, however, writes the output of each MapReduce iteration to disk. This unnecessary disk IO forms a bottleneck that results in the longer run-times. {\sc Spark}, on the other hand, can cache the intermediate results in memory, virtually eliminating the disk IO bottleneck on smaller datasets.

@@

\subsection*{Variant{\sc Spark} allows genome wide sampling of variants to improve clustering quality}
Add the results for the scalling from 1\% to X\%

>>>>

Clustering the samples 30\% fastest in \variantSpark\ compared to ADAM (1min and 20secs compared to 1min and 52secs) and  %80 vs 112
90\% faster than Hadoop (14mins and 23secs), see Table~\ref{fivewaycomparison}.
The similarity between \variantSpark\ and ADAM is expected, as they both use {\sc Spark}'s MLlib k-means implementation. 
The 30 second speed increase in \variantSpark\ is likely due us converting the VCF files to sparse vectors, whereas ADAM creates dense vectors.
%Due to the sparse nature of these row-vectors, the sparse vector format is likely to reduce some of the initial overhead.
We also compared the k-means clustering times to a multi-threaded approach in Python and R.

Although an order of magnitude slower than \variantSpark\ and ADAM, k-means clustering in both Python and R were faster than our Hadoop implementation. Python required 11min and 29sec, whereas R required 7min and 25sec. 
This demonstrates one of Hadoop's disadvantages (one which {\sc Spark} dramatically improves on): the cost of disk IO. Where Hadoop's advantage is that it can scale to datasets that exceed memory size, in this example, the entire dataset can fit in memory on a single machine.
So, where the k-means implemtations we use in R and Python store the output from each k-means iteration in memory, Hadoop writes the output of each iteration to disk, whether or not the amount of available memory is sufficient to store this data.

This unnecessary disk IO forms a bottleneck that results in the longer run-times. {\sc Spark}, on the other hand, can cache the intermediate results in memory, virtually eliminating the disk IO bottleneck on smaller datasets.

@@

\subsection*{{\sc VariantSpark} allows genome wide sampling of variants to improve clustering quality}

To demonstrate the scalability of {\sc Spark}, we continue to cluster individuals from the 1000 Genome Project, however, rather than clustering on a single chromosome, we cluster the individuals on their variants from all 22 autosomes (.
Although this pre-processing and clustering is successful, the memory required for clustering increases dramatically.

Where our previous clustering of chromosome 22 required no more than 4GB memory per executor, clustering the entire genome increases this requirement to over 20GB. On the other hand, the memory requirements for our pre-processing implementation remain unchanged, at 2 gigabytes.
One factor which results in these differing memory requirements is the sparsity of the vectors used to store the variants. Due to their sparse nature (the abundance of 0s) we can store them efficiently as Sparse Vectors, rather than Dense Vectors. However, to cluster these vectors, the k-means algorithm converts them to Dense Vectors.
With Dense Vectors, all the 0 values (which are excluded from Sparse Vectors so don't require any memory) are explicitly defined, considerably increasing the memory required.
